Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-09 manuscript.tex
3

In this section, we will review current evidence of taking advantage of spiking motifs in neurobiology, that is, from actual recordings in biological tissues. In most generic computational models, the neural activity is assumed to be encoded in the firing rate. For instance, the output of the so-called linear non-linear (L-NL) models is assumed to model the response of a biological neuron as the sequence of a linear integration followed by a non-linear spiking response generating spikes according to a Poisson point-process. As such, a simple decoding strategy is to infer the input knowing the neuron's tuning curves~\citep{jazayeri_optimal_2006} or simply by a simple regression~\citep{berens_fast_2012}. This latter model assumes a Bernoulli model for the generation of spikes, such that the decoding amounts to a single-layer logistic regression. An important perspective of these methods used to decipher the recorded activity is that it could be ultimately used to fit neural network models to the recorded activity~\citep{bellec_fitting_2021}. In that paper, authors fit summary statistics of neural data with a differentiable spiking network model. The loss function is the cross entropy (following Bernouilli hypothesis with a GLM where each unit is modelled with an SRM neuron~\citep{gerstner_time_1995} with recurrent dynamics). In particular, it comes with \href{https://github.com/EPFL-LCN/pub-bellec-wang-2021-sample-and-measure}{code} and uses the publicly available \href{http://crcns.org/data-sets/vc/pvc-11}{V1-dataset} which allows supervising the model with the input being the movie and the output the spikes recorded. These type of model may infer sparse activity in a set of binary latent variables, each describing the activity of a cell assembly~\citep{warner_probabilistic_2022}.
Using the right metric, as implemented in that paper by the chosen neural models, is essential in better understanding neural data. As such, all these models are dependent on a core definition of spike measures, and we will review here how precise spiking motifs are taken into account by such spike distances.

@@

There are different solutions to provide with a distance between two given spike trains. A known measure is the Victor-Purpura distance, which showed that their findings are overcoming inconsistencies experienced with Poisson models of spike trains~\citep{victor_nature_1996}. In a later study,~\citet{van_rossum_novel_2001} have defined a distance that uses a time constant as a parameter. Depending on this parameter, the distance interpolates between a coincidence detector and a rate difference counter. Such distances were extended to non-Euclidean metrics and use morphological manipulations to compute spike train dissimilarity~\citet{kreuz_measuring_2007}. Mathematically, the stability of distance measures induced by level-crossing sampling can be evaluated~\citep{moser_stability_2014}, notably in light of the so-called Weyl's discrepancy measure~\citep{weyl_ueber_1916}, which may lead to the definition of a cross-correlation measure which is adapted to the event-based nature of spiking signals. These observations lead to the intuition that each distance may be related as the optimal solution of a generative model for these measures, possibly through non-linear relations~\citep{aronov_non-euclidean_2004}. % ~\citep{}

@@

Another important algorithm, called SPOTDisClust, is based on the detection of structured temporal patterns~\citep{grossberger_unsupervised_2018}. They introduced an unsupervised method based for their detection from high-dimensional neural ensembles. The algorithm measures similarity between two ensemble spike patterns by determining the minimum transport cost of transforming their corresponding normalized cross-correlation matrices into each other. Many approaches to this problem are supervised. In other words, they take patterns occurring concurrently with a known event, such as the delivery of a stimulus for sensory neurons or the traversal of a running track for hippocampal place fields, as a ``template'' and then search for repetitions of the same template in spiking activity~\citep{nadasdy_replay_1999,lee_combinatorial_2004}. In SPOTDisClust, the learning is unsupervised. It uses the prior that there is only one spike per pattern. Using a so-called t-SNE projection (that allows to project this high-dimensional representation to a lower-dimension map) validated that this clustering method can retrieve all patterns from the data (see Figure~\ref{fig:SPOTDisClust}). The limits of this method are that it is computationally complex, block-based and strictly specialized for the task at hand. To overcome these difficulties, a novel method was recently developed~\citep{sotomayor-gomez_spikeship_2021}.
%   Davidson TJ, Kloosterman F, Wilson MA. Hippocampal replay of extended experience. Neuron. 2009;63(4):497--507. pmid:19709631

@@

% \subsection{Extending the decoding at multiple time scales}
In the paper by~\citet{russo_cell_2017}, the authors present a unifying methodological and conceptual framework which detects assembly structure at many different time scales, levels of precision, and with arbitrary internal organization. It uses  sliding window as in~\citep{grun_unitary_2002} and the reliable and efficient analysis of an excess or deficiency of joint-spike events~\citep{pipa_neuroxidence_2008}. They extend the measure to multiple lags~\citep{torre_synchronous_2016}. The core measure is based on a non-stationarity corrected parametric test statistic for assessing the independence of pairs and an agglomerative, heuristic clustering algorithm for fusing significant pairs into higher-order assemblies.  To overcome the limits of models which either require spike times to be discretized, utilize a suboptimal least-squares criterion, or do not provide uncertainty estimates for model predictions or estimated parameters,~\citep{williams_point_2020} address each of these shortcomings by developing a point process model that characterizes fine-scale sequences at the level of individual spikes and represents sequence occurrences as a few marked events in continuous time. They also introduce learnable time warping parameters to model sequences of varying duration, which have been experimentally observed in neural circuits, and demonstrate these advantages on experimental recordings from songbird higher vocal center and rodent hippocampus.
At a larger scale, in~\citep{van_kempen_top-down_2021}, it was shown that attentional information from V4 or arousal can change the timings of groups of events in V1. They develop a Hidden Markov Model for quantifying the transitions. In particular, they show that fluctuations in neural excitability are coordinated between visual areas with retinotopic precision. Top-down attention drives inter-areal coordination along the reverse cortical hierarchy, predicting better behavioral performance with increased coordination. Building such models for predicting changes of timings based on context, such as using a so-called change point model for blocked-based experimental protocols~\citep{pasturel_humans_2020}, could therefore provide useful prior information to enhance the decoding from neural activity.

@@

\subsection{What biological mechanism enables delay learning?}

Despite the extensive computer evidence of the effectiveness of precise spiking detection we presented above, doubts may remain as to the reliability of this learning mechanism and whether there is a real need for further research on this subject. The discovery of the existence of an equivalent biological mechanism in Vertebrates as well as the demonstration of the importance of its role in various developed behaviours allow us to put these doubts to rest. In the following paragraphs we will successively describe the first biological observations of biological delay learning, identify myelination as an important actor and finally study this phenomenon at the molecular level.

>>>>

In this section, we will review current evidence of taking advantage of spiking motifs in neurobiology, that is, from actual recordings in biological tissues. In most generic computational models, the neural activity is assumed to be encoded in the firing rate. For instance, the output of the so-called linear non-linear (L-NL) models is assumed to model the response of a biological neuron as the sequence of a linear integration followed by a non-linear spiking response generating spikes according to a Poisson point-process. As such, a simple decoding strategy is to infer the input knowing the neuron's tuning curves~\citep{jazayeri_optimal_2006} or simply by a simple regression~\citep{berens_fast_2012}. This latter model assumes a Bernoulli model for the generation of spikes, such that the decoding amounts to a single-layer logistic regression. An important perspective of these methods used to decipher the recorded activity is that it could be ultimately used to fit neural network models to the recorded activity~\citep{bellec_fitting_2021}. In this particular paper, authors fit summary statistics of neural data with a (differentiable) spiking network model. The loss function is the cross entropy (following Bernouilli hypothesis with a GLM where each unit is modelled with an SRM neuron~\citep{gerstner_time_1995}) and embedded with recurrent dynamics. In particular, it comes with \href{https://github.com/EPFL-LCN/pub-bellec-wang-2021-sample-and-measure}{code} and uses the publicly available \href{http://crcns.org/data-sets/vc/pvc-11}{V1-dataset} which allows supervising the model with the input being the movie and the output the spikes recorded. These type of model may infer sparse activity in a set of binary latent variables, each describing the activity of a cell assembly~\citep{warner_probabilistic_2022}.
Using the right metric, as implemented in that paper by the corresponding neural models, is essential in better understanding neural data. Importantly, these models are dependent on a core definition of spike measures, and we will review here how precise spiking motifs are taken into account by such spike distances.

@@

There are different solutions to provide with a distance between two given spike trains. A known measure is the Victor-Purpura distance, which  overcomes inconsistencies experienced with a simple firing rate (Poisson model) of spike trains~\citep{victor_nature_1996}. In a later study,~\citet{van_rossum_novel_2001} have defined a distance that uses a time constant as a parameter. Depending on this parameter, the distance interpolates between a coincidence detector and a rate difference counter. Such distances were extended to non-Euclidean metrics and use morphological manipulations to compute spike train dissimilarity~\citep{kreuz_measuring_2007}. Mathematically, the stability of distance measures induced by level-crossing sampling can be evaluated~\citep{moser_stability_2014}, notably in light of the so-called Weyl's discrepancy measure~\citep{weyl_ueber_1916}, which may lead to the definition of a cross-correlation measure which is adapted to the event-based nature of spiking signals. These observations lead to the intuition that each distance may be related as the optimal solution of a generative model for these measures, possibly through non-linear relations~\citep{aronov_non-euclidean_2004}. % ~\citep{}

@@

Another important algorithm, called SPOTDisClust, is based on the detection of structured temporal patterns~\citep{grossberger_unsupervised_2018}. They introduced an unsupervised method based for their detection from high-dimensional neural ensembles. The algorithm measures similarity between two ensemble spike patterns by determining the minimum transport cost of transforming their corresponding normalized cross-correlation matrices into each other. Many approaches to this problem are supervised. In other words, they take patterns occurring concurrently with a known event, such as the delivery of a stimulus for sensory neurons or the traversal of a running track for determining hippocampal place fields, as a ``template'' and then search for repetitions of the same template in spiking activity~\citep{nadasdy_replay_1999,lee_combinatorial_2004}. In SPOTDisClust, the learning is unsupervised. It uses the prior that there is only one spike per pattern. Using a so-called t-SNE projection (that allows to project this high-dimensional representation to a lower-dimension map) validated that this clustering method can retrieve all patterns from the data (see Figure~\ref{fig:SPOTDisClust}). The limits of this method are that it is computationally complex, block-based and strictly specialized for the task at hand. To overcome these difficulties, a novel method was recently developed~\citep{sotomayor-gomez_spikeship_2021}.
%   Davidson TJ, Kloosterman F, Wilson MA. Hippocampal replay of extended experience. Neuron. 2009;63(4):497--507. pmid:19709631

@@

% \subsection{Extending the decoding at multiple time scales}
In the paper by~\citet{russo_cell_2017}, the authors present a unifying methodological and conceptual framework which detects assembly structure at many different time scales, levels of precision, and with arbitrary internal organization. It uses  sliding window as in~\citep{grun_unitary_2002} and the reliable and efficient analysis of an excess or deficiency of joint-spike events~\citep{pipa_neuroxidence_2008}. They extend the measure to multiple lags~\citep{torre_synchronous_2016}. The core measure is based on a non-stationarity corrected parametric test statistic for assessing the independence of pairs and an agglomerative, heuristic clustering algorithm for fusing significant pairs into higher-order assemblies.  To overcome the limits of models which either require spike times to be discretized, utilize a suboptimal least-squares criterion, or do not provide uncertainty estimates for model predictions or estimated parameters,~\citep{williams_point_2020} address each of these shortcomings by developing a point process model that characterizes fine-scale sequences at the level of individual spikes and represents sequence occurrences as a few marked events in continuous time. They also introduce learnable time-warping parameters to model sequences of varying duration, which have been experimentally observed in neural circuits, and demonstrate these advantages on experimental recordings from songbird higher vocal center and rodent hippocampus.
At a larger scale, in~\citep{van_kempen_top-down_2021}, it was shown that attentional information from V4 or arousal can change the timings of groups of events in V1. They develop a Hidden Markov Model for quantifying the transitions. In particular, they show that fluctuations in neural excitability are coordinated between visual areas with retinotopic precision. Top-down attention drives inter-areal coordination along the reverse cortical hierarchy, predicting better behavioral performance with increased coordination. Building such models for predicting changes of timings based on context, such as using a so-called change point model for blocked-based experimental protocols~\citep{pasturel_humans_2020}, could therefore provide useful prior information to enhance the decoding from neural activity.

@@

\subsection{What biological mechanism could allow learning spiking motifs?}

Despite the extensive computer evidence of the effectiveness of precise spiking detection we presented above, doubts may remain as to the reliability of this learning mechanism and whether there is a real need for further research on this subject. The discovery of the existence of an equivalent biological mechanism in vertebrates as well as the demonstration of the importance of its role in various developed behaviours allow us to put these doubts to rest. In the following paragraphs we will successively describe the first biological observations of biological delay learning, identify myelination as an important actor and finally study this phenomenon at the molecular level.
